---
title: "Data Wrangling"
author: "Martin Olarte"
date: "2023-03-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Web Scraping - NBA Combine Anthro Data

```{r}
library(tidyverse)
library(rvest)
library(readr)

create_anthro_df <- function(season_start_year) {
  html_tbl <- read_file(paste0("html_tables/",season_start_year,"_", season_start_year+1))

  df <- as.data.frame(read_html(html_tbl) %>% 
                        html_table(fill = TRUE))
  anthro_df <- df %>%
    mutate(BODY.FAT.. = as.numeric(str_remove(BODY.FAT.., "%")),
           HAND.LENGTH..inches. = as.numeric(HAND.LENGTH..inches.) * 2.54,
           HAND.WIDTH..inches. = as.numeric(HAND.WIDTH..inches.) * 2.54,
           HEIGHT.W.O.SHOES = as.numeric(str_extract(HEIGHT.W.O.SHOES, "\\d+")) * 30.48 + as.numeric(str_extract(gsub(" ", "", HEIGHT.W.O.SHOES), "(?<=')[^']+(?=')")) * 2.54,
           HEIGHT.W..SHOES = as.numeric(str_extract(HEIGHT.W..SHOES, "\\d+")) * 30.48 + as.numeric(str_extract(gsub(" ", "", HEIGHT.W..SHOES), "(?<=')[^']+(?=')")) * 2.54,
           STANDING.REACH = as.numeric(str_extract(STANDING.REACH, "\\d+")) * 30.48 + as.numeric(str_extract(gsub(" ", "", STANDING.REACH), "(?<=')[^']+(?=')")) * 2.54,
           WEIGHT..LBS. = as.numeric(WEIGHT..LBS.),
           WINGSPAN = as.numeric(str_extract(WINGSPAN, "\\d+")) * 30.48 + as.numeric(str_extract(gsub(" ", "", WINGSPAN), "(?<=')[^']+(?=')")) * 2.54,
           SEASON_START_YEAR = 2000 + season_start_year)
  
  return(anthro_df)
}

all_anthro <- suppressWarnings(lapply(0:22, create_anthro_df))

all_anthro_df <- do.call(rbind, all_anthro)

all_anthro_df <- all_anthro_df %>%
  select(PLAYER, SEASON_START_YEAR, everything())
```

## Download Basketball Reference (BR) Data

Rk -- Rank
Pos -- Position
Age -- Player's age on February 1 of the season
Tm -- Team
G -- Games
MP -- Minutes Played
PER -- Player Efficiency Rating
A measure of per-minute production standardized such that the league average is 15.
TS% -- True Shooting Percentage
A measure of shooting efficiency that takes into account 2-point field goals, 3-point field goals, and free throws.
3PAr -- 3-Point Attempt Rate
Percentage of FG Attempts from 3-Point Range
FTr -- Free Throw Attempt Rate
Number of FT Attempts Per FG Attempt
ORB% -- Offensive Rebound Percentage
An estimate of the percentage of available offensive rebounds a player grabbed while they were on the floor.
DRB% -- Defensive Rebound Percentage
An estimate of the percentage of available defensive rebounds a player grabbed while they were on the floor.
TRB% -- Total Rebound Percentage
An estimate of the percentage of available rebounds a player grabbed while they were on the floor.
AST% -- Assist Percentage
An estimate of the percentage of teammate field goals a player assisted while they were on the floor.
STL% -- Steal Percentage
An estimate of the percentage of opponent possessions that end with a steal by the player while they were on the floor.
BLK% -- Block Percentage
An estimate of the percentage of opponent two-point field goal attempts blocked by the player while they were on the floor.
TOV% -- Turnover Percentage
An estimate of turnovers committed per 100 plays.
USG% -- Usage Percentage
An estimate of the percentage of team plays used by a player while they were on the floor.
OWS -- Offensive Win Shares
An estimate of the number of wins contributed by a player due to offense.
DWS -- Defensive Win Shares
An estimate of the number of wins contributed by a player due to defense.
WS -- Win Shares
An estimate of the number of wins contributed by a player.
WS/48 -- Win Shares Per 48 Minutes
An estimate of the number of wins contributed by a player per 48 minutes (league average is approximately .100)
OBPM -- Offensive Box Plus/Minus
A box score estimate of the offensive points per 100 possessions a player contributed above a league-average player, translated to an average team.
DBPM -- Defensive Box Plus/Minus
A box score estimate of the defensive points per 100 possessions a player contributed above a league-average player, translated to an average team.
BPM -- Box Plus/Minus
A box score estimate of the points per 100 possessions a player contributed above a league-average player, translated to an average team.
VORP -- Value over Replacement Player
A box score estimate of the points per 100 TEAM possessions that a player contributed above a replacement-level (-2.0) player, translated to an average team and prorated to an 82-game season. Multiply by 2.70 to convert to wins over replacement.

```{r}
# Variable names and descriptions can be found above ^
create_advanced_stats_df <- function(season_last_year) {
  url <- paste0("https://www.basketball-reference.com/leagues/NBA_", season_last_year,"_advanced.html")
  df <- as.data.frame(read_html(url) %>% 
                        html_table(fill = TRUE))
  
  advanced_stats_df <- df %>%
    select(-c(Rk, Var.20, Var.25)) %>%
    filter(Player != "Player") %>%
    mutate(SEASON_START_YEAR = season_last_year - 1) %>%
    mutate(across(!Player & !Pos & !Tm, as.numeric))
  
  return(advanced_stats_df)
}

all_advanced_stats <- suppressWarnings(lapply(2001:2023, create_advanced_stats_df))

all_advanced_stats_df <- do.call(rbind, all_advanced_stats)

# There are 13892 records
# Player + Tm (Team) is the unique key

all_advanced_stats_df <- all_advanced_stats_df %>%
  select(Player, SEASON_START_YEAR, everything())
```

## Download Kaggle Draft Data

```{r}
kaggle <- read_csv("all_seasons.csv")

draft_df <- kaggle %>%
  select(-c(...1, )) %>%
  select(player_name, season, everything()) %>%
  mutate(season = as.numeric(str_extract(season, "\\d+")),
         draft_year = as.numeric(draft_year),
         draft_round = as.numeric(draft_round),
         draft_number = as.numeric(draft_number)) %>%
  filter(!is.na(draft_round)) #filters 2274 Undrafted observations for 761 players

kaggle %>%
  filter(draft_round == "Undrafted") %>%
  group_by(player_name) %>%
  summarise(total = n())
```

## Merging All Datasets

```{r}
# PLAYER is the unique key
all_anthro_df <- read_csv("WrangledData/all_anthro_df.csv")

# 1,480 Players had one entry in the draft combine
# 36 Players had two entries in the draft combine
two_entries <- all_anthro_df %>%
  group_by(PLAYER) %>%
  summarise(total = n()) %>%
  filter(total > 1) %>%
  select(PLAYER) %>%
  pull()

# Can keep only the last instance a player went to combine, since the only way they redo the combine is if they did not get drafted
all_anthro_df <- all_anthro_df %>%
  arrange(PLAYER) %>%
  group_by(PLAYER) %>%
  slice(n())
```

```{r}
# PLAYER + Tm + SEASON_START_YEAR is the unique key
all_advanced_stats_df <- read_csv("WrangledData/all_advanced_stats_df.csv")

# 13,892 records when grouped by Player, Tm, SEASON_START_YEAR
all_advanced_stats_df %>%
  group_by(Player, Tm, SEASON_START_YEAR) %>%
  summarise(total = n()) %>%
  filter(total > 1)

# 2,261 distinct players
all_advanced_stats_df %>%
  group_by(Player)
```

```{r}
# player_name, team_abbreviation, season is the unique key
draft_df <- read_csv("WrangledData/draft_df_kaggle.csv")

draft_df <- draft_df %>%
  filter(season >= 2000)

# 8,536 records when grouped by player_name, team_abbreviation, season
draft_df %>%
  group_by(player_name, team_abbreviation, season) %>%
  summarise(total = n()) %>%
  filter(total > 1)

# 1,509 distinct players
draft_df %>%
  group_by(player_name)

draft_df %>%
  group_by(draft_number) %>%
  summarise(total = n())

# Manually checking draft picks beyond the two normal rounds (60 picks)
draft_df %>%
  filter(draft_number > 60)

# All other records are correct (double checked online)

# Walker Russell refers to Walker Russell Jr. in stats, but draft information is from Walker Russell Sr.
draft_df %>%
  filter(player_name == "Walker Russell")

# Manually removing him since he went undrafted (https://en.wikipedia.org/wiki/Walker_Russell_Jr.)
draft_df <- draft_df %>%
  filter(player_name != "Walker Russell")

##### RE RUNNING ANALYSIS

# 8,535 records when grouped by player_name, team_abbreviation, season
# 1,508 distinct players
draft_df %>%
  group_by(player_name)
```
```{r}
# PLAYER is the unique key
# 1,516 players
all_anthro_df %>%
  arrange(SEASON_START_YEAR, PLAYER)

# PLAYER + Tm + SEASON_START_YEAR is the unique key
# 13,892 records when grouped by Player, Tm, SEASON_START_YEAR
# 2,261 distinct players
all_advanced_stats_df %>%
  arrange(SEASON_START_YEAR, Player)


# player_name is the unique key
# 1,508 distinct players
draft_df <- draft_df %>%
  group_by(player_name) %>%
  slice(n()) %>%
  select(player_name, college, country, draft_year, draft_round, draft_number)

drafted_players <- draft_df %>%
  select(player_name) %>%
  pull()

all_advanced_stats_df <- all_advanced_stats_df %>%
  filter(Player %in% drafted_players)
  

advanced_stats_drafted_df <- left_join(all_advanced_stats_df, draft_df, by = c("Player" = "player_name"))

advanced_stats_drafted_df <- advanced_stats_drafted_df %>% 
  arrange(Player, SEASON_START_YEAR)

full_tbl <- left_join(advanced_stats_drafted_df, all_anthro_df, by = c("Player" = "PLAYER"))

# Create a vector of new column names with proper capitalization and informative names
new_col_names <- c("Player Name", "Season Start Year", "Position", "Age", "Team", "Games Played", "Minutes Played", 
                   "Player Efficiency Rating", "True Shooting Percentage", "3-Point Attempt Rate", "Free Throw Rate", 
                   "Offensive Rebound Percentage", "Defensive Rebound Percentage", "Total Rebound Percentage", 
                   "Assist Percentage", "Steal Percentage", "Block Percentage", "Turnover Percentage", 
                   "Usage Percentage", "Offensive Win Shares", "Defensive Win Shares", "Total Win Shares", 
                   "Win Shares Per 48 Minutes", "Offensive Box Plus/Minus", "Defensive Box Plus/Minus", "Box Plus/Minus", 
                   "Value Over Replacement Player", "College", "Country", "Draft Year", "Draft Round", "Draft Number", 
                   "Combine Year", "Combine Position","Body Fat Percentage", "Hand Length (cm)", "Hand Width (cm)", 
                   "Height Without Shoes (cm)", "Height With Shoes (cm)", "Standing Reach (cm)", 
                   "Weight (lbs)", "Wingspan")

# Rename the columns in the original dataframe with the new column names
colnames(full_tbl) <- new_col_names
```

